Staged Compilation With Two-Level Type Theory

### 2022-07-20 paper/paper.tex
camera ready edit

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.

%% \setcopyright{acmcopyright}
%% \copyrightyear{2018}
%% \acmYear{2018}
%% \acmDOI{XXXXXXX.XXXXXXX}

@@

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

@@

\title{Staged Compilation With Two-Level Type Theory}

@@

\begin{CCSXML}
\end{CCSXML}

\keywords{type theory, two-level type theory, staged compilation}
\maketitle

@@

\emph{Two-level type theory}, or 2LTT in short, was described by Annekov,
Capriotti, Kraus and Sattler \cite{twolevel}, building on ideas from Vladimir
Voevodsky \cite{hts}. The motivation was to allow convenient metatheoretical
reasoning about a certain mathematical language (homotopy type theory), and to
enable concise and modular ways to extend the language with axioms.

@@

\subsubsection{Moving between stages}
At this point, our system is rather limited, since there is no interaction
between the stages. We enable such interaction through the following operations.
\begin{itemize}
\item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
  staging point of view, $\Lift A$ is the type of metaprograms which compute
  runtime expressions of type $A$.
\item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
  A quoted term $\qut{t}$ represents the metaprogram which immediately yields
  $t$.
\item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
  During staging, the metaprogram in the splice is executed, and the resulting
  expression is inserted into the output.

@@

\item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
  $\qut{\spl t}=t$ as definitional equalities.
\end{itemize}
Note that none of these three operations can be expressed as functions, since
function types cannot cross between stages.

@@

This is fairly similar to the well-known monomorphization models in the C++ and
Rust programming languages. The evident difference is that in 2LTT we can use a
full type theory at compile time. Additionally, we will see in Section
\ref{sec:intensional-analysis} that this setup is compatible with intensional
analysis, the ability to analyze the structure of object types and terms.

@@

\subsubsection{Metatheory}
The metatheory is an intensional type theory with a cumulative universe
hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
1$. The transfinite $i$ levels are a convenience feature for modeling typing
contexts, where sometimes we will need universes large enough to fit
types at arbitrary finite levels. We also assume uniqueness of identity proofs
and function extensionality. We will omit transports along identity proofs in
the paper, for the sake of readability. The missing transports can be in
principle recovered due to Hofmann's conservativity result
\cite{hofmann95extensional}; see also Winterhalter et
al. \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$, $\top$ and
$\mbb{N}$ types in all universes, and reuse Notation \ref{notation:basics}. We
also assume certain quotient inductive-inductive types, to represent the syntax of 2LTT
and the object theory. We will detail these shortly in Sections \ref{sec:2lttmod}
and \ref{sec:objmod}.

@@

Fortunately, our specified meta type theory can be interpreted in
\emph{setoids}, as most recently explained by Pujet and Tabareau
\cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
equivalence relation on it. We interpret closed types as setoids, and
dependent types as setoid fibrations. For inductive base types such $\Bool$ or
$\Nat$, the equivalence relation is interpreted as identity. Functions are
interpreted as equivalence-preserving functions, and two functions are equivalent if
they are pointwise equivalent. A quotient type $A/R$ is interpreted by extending
the semantic notion of equivalence for $A$ with the given $R$ relation. We will
sketch the interpretation of the 2LTT syntax (given as a quotient inductive-inductive type)
in Section \ref{sec:2lttsetoid}.

@@

In other words, we can view the usage of constructive quotients as a convenient
shorthand for working with explicit equivalence relations. When we reason about
the syntax of a type theory, this provides a bridge between the
categorical/algebraic style and the traditional style that uses separate
equivalence relations. It makes sense to work with quotients internally to a
type theory whenever we only want to talk about constructions and properties
which respect conversion. However, at any point we have the option to switch to
an external perspective and look at underlying sets and the underlying maps.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parametrized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-level fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

@@

\subsubsection{Meta-level family structure}
We interpret context extension and type/term substitution as follows. Note the
usage of the pattern matching notation on the $\rexti$ isomorphism.
\begin{alignat*}{4}
  & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
    (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
  & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
  & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
\end{alignat*}
It is enough to specify the $\blank\rel$ action on extended substitutions
$(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
\ext A)\,(A[\p])$. The category-with-families equations hold evidently.
\begin{alignat*}{4}
  & (\sigma,\,t)\rel\,\gamma\rel                 && := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel)\\
  & \p\rel\,(\gamma\rel,\,t\rel)                 && := \gamma\rel\\
  & \q\rel\,(\gamma\rel,\,t\rel)                 && := t\rel
\end{alignat*}

@@

The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
the thesis of Capriotti \cite{capriotti2017models} and ideas from Voevodsky
\cite{hts}, is the foremost inspiration in the current work. Interestingly, the
above works do mention metaprogramming as a source of intuition for 2LTT, but
they only briefly touch on this aspect, and in \cite{twolevel} the main focus is
on extensions of basic 2LTT which do not have staging semantics
anymore. Ibid.\ conjectures the strong conservativity of 2LTT in Proposition
2.18, and suggests categorical gluing as proof strategy. In this work, we do
essentially that: our presheaf model and the logical relation model could be
merged into a single instance of gluing along the $\re$ morphism; this merges
staging and its soundness proof. We split this construction to two parts in order
to get a self-contained presentation of the staging algorithm.

The multi-stage calculus of Kawata and Igarashi \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have large elimination,
hence it does not support computing types by staging. Second, it does not
support staged compilation in the sense of Definition \ref{def:staging}; rather,
it is a system for runtime code manipulation and evaluation (which we may call
staged \emph{computation}). In other words, programs may manipulate code
fragments during execution, in more liberal ways than in 2LTT, but there is no
judgment which ensures that a purely ``object-level'' program can be extracted
from a multi-stage program. Similar considerations apply to systems with the S4
modality (see e.g.\ Davies and Pfenning \cite{DBLP:journals/jacm/DaviesP01}),
where there is no judgment that guarantees elimination of modal operations.

@@

\bibliographystyle{ACM-Reference-Format}
\interlinepenalty=10000
\bibliography{references}

>>>>

%%% The following is specific to ICFP '22 and the paper
%%% 'Staged Compilation with Two-Level Type Theory'
%%% by András Kovács.
%%%
\setcopyright{rightsretained}
\acmPrice{}
\acmDOI{10.1145/3547641}
\acmYear{2022}
\copyrightyear{2022}
\acmSubmissionID{icfp22main-p52-p}
\acmJournal{PACMPL}
\acmVolume{6}
\acmNumber{ICFP}
\acmArticle{110}
\acmMonth{8}

@@

\bibliographystyle{ACM-Reference-Format}
%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
\citestyle{acmauthoryear}

@@

\title{Staged Compilation With Two-Level Type Theory}
\titlenote{The author was supported by the “Application Domain Specific Highly
  Reliable IT Solutions” project which has been implemented with support from
  the National Research, Development and Innovation Fund of Hungary, financed
  under the Thematic Excellence Programme TKP2020-NKA-06 (National Challenges
  Subprogramme) funding scheme.}

@@

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10011047</concept_id>
       <concept_desc>Software and its engineering~Source code generation</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Software and its engineering~Source code generation}

\keywords{type theory, two-level type theory, staged compilation}

\maketitle

@@

\emph{Two-level type theory}, or 2LTT in short, was described in
\cite{twolevel}, building on ideas from \cite{hts}. The motivation was to allow
convenient metatheoretical reasoning about a certain mathematical language
(homotopy type theory), and to enable concise and modular ways to extend the
language with axioms.

@@

\subsubsection{Moving between stages}
At this point, our system is rather limited, since there is no interaction
between the stages. We enable such interaction through the following operations.
Note that none of these operations can be expressed as functions, since function
types cannot cross between stages.
\begin{itemize}
\item \emph{Lifting:} for $A : \U_0$, we have $\Lift A : \U_1$.  From the
  staging point of view, $\Lift A$ is the type of metaprograms which compute
  runtime expressions of type $A$.
\item \emph{Quoting:} for $A : \U_0$ and $t : A$, we have $\qut{t} :\,\Lift A$.
  A quoted term $\qut{t}$ represents the metaprogram which immediately yields
  $t$.
\item \emph{Splicing:} for $A : \U_0$ and $t :\,\Lift A$, we have $\spl t : A$.
  During staging, the metaprogram in the splice is executed, and the resulting
  expression is inserted into the output.

@@

\item Quoting and splicing are definitional inverses, i.e.\ we have $\spl\qut{t} = t$ and
  $\qut{\spl t}=t$ as definitional equalities.
\end{itemize}

@@

%% This is fairly similar to the well-known monomorphization models in the C++ and
%% Rust programming languages. The evident difference is that in 2LTT we can use a
%% full type theory at compile time. Additionally, we will see in Section
%% \ref{sec:intensional-analysis} that this setup is compatible with intensional
%% analysis, the ability to analyze the structure of object types and terms.

@@

\subsubsection{Metatheory}
The metatheory is an intensional type theory with a cumulative universe
hierarchy $\Set_i$, where $i$ is ordinal number such that $i \leq \omega +
1$. The transfinite $i$ levels are a convenience feature for modeling typing
contexts, where sometimes we will need universes large enough to fit types at
arbitrary finite levels. We also assume uniqueness of identity proofs and
function extensionality. We will omit transports along identity proofs in the
paper, for the sake of readability. The missing transports can be in principle
recovered due to Hofmann's conservativity result \cite{hofmann95extensional};
see also \cite{DBLP:conf/cpp/WinterhalterST19}. We assume $\Pi$, $\Sigma$,
$\top$ and $\mbb{N}$ types in all universes, and reuse Notation
\ref{notation:basics}. We also assume certain quotient inductive-inductive
types, to represent the syntax of 2LTT and the object theory. We will detail
these shortly in Sections \ref{sec:2lttmod} and \ref{sec:objmod}.

@@

Fortunately, our specified meta type theory can be interpreted in
\emph{setoids}, as most recently explained in
\cite{DBLP:journals/pacmpl/PujetT22}. A setoid is a set together with an
equivalence relation on it. We interpret closed types as setoids, and dependent
types as setoid fibrations. For inductive base types such $\Bool$ or $\Nat$, the
equivalence relation is interpreted as identity. Functions are interpreted as
equivalence-preserving functions, and two functions are equivalent if they are
pointwise equivalent. A quotient type $A/R$ is interpreted by extending the
semantic notion of equivalence for $A$ with the given $R$ relation. We will
sketch the interpretation of the 2LTT syntax (given as a quotient
inductive-inductive type) in Section \ref{sec:2lttsetoid}.

@@

In other words, we can view quotients as a convenient shorthand for working with
explicit equivalence relations. We can work with quotients when we want to enforce
preservation of equivalence relations, but at any point we have the option to
switch to an external view and look at underlying sets and functions.

%% In other words, we can view the usage of constructive quotients as a convenient
%% shorthand for working with explicit equivalence relations. When we reason about
%% the syntax of a type theory, this provides a bridge between the
%% categorical/algebraic style and the traditional style that uses separate
%% equivalence relations.
%% It makes sense to work with quotients internally to a type theory whenever we
%% only want to talk about constructions and properties which respect
%% conversion. However, at any point we have the option to switch to an external
%% perspective and look at underlying sets and the underlying maps.

@@

\noindent The basic scheme for internalization is as follows:
\begin{alignat*}{3}
  & \Gamma : \hCon                  && \hspace{1em}\text{is internalized as}\hspace{1em} \Gamma : \whset_\omega\\
  & \sigma : \hSub\,\Gamma\,\Delta  && \hspace{1em}\text{is internalized as}\hspace{1em} \sigma : \Gamma \to \Delta\\
  & A : \hTy_{1,j}\,\Gamma           && \hspace{1em}\text{is internalized as}\hspace{1em} A      : \Gamma \to \whset_j \\
  & t : \hTm_{1,j}\,\Gamma\,A        && \hspace{1em}\text{is internalized as}\hspace{1em} t      : (\gamma : \Gamma) \to A\,\gamma
\end{alignat*}
\subsubsection{Object-theoretic syntax} The syntax of the object theory is clearly
fully stable under object-theoretic substitution, so we can internalize all of
its type and term formers. We internalize object-level types and terms as
$\Ty_{\mbbo j} : \whset_0$ and $\Tm_{\mbbo j} : \Ty_{\mbbo j} \to
\whset_0$. $\Ty_{\mbbo j}$ is closed under type formers. For instance, we have
\[ \Nat_j : \Ty_{\mbbo j}\hspace{1.5em}\zero_j : \Tm_{\mbbo j}\,\Nat_j\hspace{1.5em}
    \suc_j : \Tm_{\mbbo j}\,\Nat_j \to \Tm_{\mbbo j}\,\Nat_j \]
together with $\NatElim$, and likewise we have all other type formers.
\subsubsection{Internal $\ev$} We also use an internal view of
$\ev$ which maps 2LTT syntax to internal values; i.e.\ we compose $\ev$ with
internalization. Below, note that the input to $\ev$ is \emph{external}, so we
mark $\ev$ as being parameterized by external input.
\begin{alignat*}{3}
  &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
  &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
  &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
  &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
  &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
  &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
\end{alignat*}
\subsubsection{Object-level fragment of 2LTT}
The purely object-level syntactic fragment of 2LTT can be internalized as
follows. We define externally the presheaf of object-level types as
$|\Ty_{0,j}|\,\Gamma := \Ty_{0,j}\,\emb{\Gamma}$, and the displayed presheaf of
object-level terms over $\Ty_{0,j}$ as $|\Tm_{0,j}|\,\{\Gamma\}\,A :=
\Tm_{0,j}\,\emb{\Gamma}\,A$. Hence, internally we have $\Ty_{0,j} : \whset_0$ and
$\Tm_{0,j} : \Ty_{0,j} \to \whset_0$. $\Ty_{0,j}$ is closed under all type formers,
analogously as we have seen for $\Ty_{\mbbo j}$.

%% \begin{alignat*}{6}
%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega && \ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \hspace{3.5em}&& \ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j&& \ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
%% \end{alignat*}

%% \begin{alignat*}{3}
%%   &\ev\,(\Gamma &&: \Con)                      &&: \whset_\omega \\
%%   &\ev\,(\sigma &&: \Sub\,\Gamma\,\Delta)      &&: \ev\,\Gamma \to \ev\,\Delta \\
%%   &\ev\,(A      &&: \Ty_{1,j}\,\Gamma)          &&: \ev\,\Gamma \to \whset_j\\
%%   &\ev\,(t      &&: \Tm_{1,j}\,\Gamma)          &&: (\gamma : \ev\,\Gamma) \to \ev\,A\,\gamma\\
%%   &\ev\,(A      &&: \Ty_{0,j}\,\Gamma)          &&: \ev\,\Gamma \to \Ty_{\mbbo_j}\\
%%   &\ev\,(t      &&: \Tm_{0,j}\,\Gamma\,A)       &&: (\gamma : \ev\,\Gamma) \to \Tm_{\mbbo j}\,(\ev\,A\,\gamma)
%% \end{alignat*}

@@

\subsubsection{Meta-level family structure}
We interpret context extension and type/term substitution as follows. Note the
usage of the pattern matching notation on the $\rexti$ isomorphism.
\begin{alignat*}{4}
  & (\Gamma \ext_1 A)\rel\,(\gamma,\,\alpha)\,(\rexti(\gamma',\,\alpha')) && :=
    (\gamma\rel : \Gamma\rel\,\gamma\,\gamma') \times A\rel\,\gamma\rel\,\alpha\,\alpha'\\
  & (A[\sigma])\rel\,\gamma\rel\,\alpha\,\alpha' && := A\rel\,(\sigma\rel\,\gamma\rel)\,\alpha\,\alpha'\\
  & (t[\sigma])\rel\,\gamma\rel\                 && := t\rel\,(\sigma\rel\,\gamma\rel)
\end{alignat*}
It is enough to specify the $\blank\rel$ action on extended substitutions
$(\sigma,\,t) : \Sub\,\Gamma\,(\Delta\ext A)$, $\p : \Sub\,(\Gamma\ext
A)\,\Gamma$, $\p : \Sub\,(\Gamma\ext A)\,\Gamma$ and $\q : \Tm_{1,j}\,(\Gamma
\ext A)\,(A[\p])$. The category-with-families equations hold evidently.
\begin{alignat*}{4}
  & (\sigma,\,t)\rel\,\gamma\rel := (\sigma\rel\,\gamma\rel,\,t\rel\,\gamma\rel) \hspace{2em} \p\rel\,(\gamma\rel,\,t\rel) := \gamma\rel \hspace{2em} \q\rel\,(\gamma\rel,\,t\rel) := t\rel
\end{alignat*}

@@

The work Annekov et al.\ on two-level type theory \cite{twolevel}, building on
\cite{capriotti2017models} and \cite{hts}, is the foremost inspiration in
the current work. Interestingly, the above works do mention metaprogramming as a
source of intuition for 2LTT, but they only briefly touch on this aspect, and in
\cite{twolevel} the main focus is on extensions of basic 2LTT which do not have
staging semantics anymore. Ibid.\ conjectures the strong conservativity of 2LTT
in Proposition 2.18, and suggests categorical gluing as proof strategy. In this
work, we do essentially that: our presheaf model and the logical relation model
could be merged into a single instance of gluing along the $\re$ morphism; this
merges staging and its soundness proof. We split this construction to two parts
in order to get a self-contained presentation of the staging algorithm.

The multi-stage calculus in \cite{multi-stage-calculus}
supports staging operations and type dependencies. However, there are major
differences to the current work. First, ibid.\ does not have large elimination,
hence it does not support computing types by staging. Second, it does not
support staged compilation in the sense of Definition \ref{def:staging}; rather,
it is a system for runtime code manipulation and evaluation (which we may call
staged \emph{computation}). In other words, programs may manipulate code
fragments during execution, in more liberal ways than in 2LTT, but there is no
judgment which ensures that a purely ``object-level'' program can be extracted
from a multi-stage program. Similar considerations apply to systems with the S4
modality (see e.g.\ \cite{DBLP:journals/jacm/DaviesP01}),
where there is no judgment that guarantees elimination of modal operations.

@@

\interlinepenalty=10000
\bibliography{references}
